---
title: "Invertebrate catch to CPUE from AKFIN"
author: "Alberto Rovellini"
date: "13/11/2021"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

This document takes RACE-GAP bottom trawl catch data from AKFIN, size-composition information, size at 50% maturity, and length-weight relationships to:

1. Calculate CPUE at haul level based on catch and haul information. We do not use CPUE data from AKFIN because that is not available for all species (e.g., not for invertebrates).
2. Based on size-composition data and length at 50% maturity, label each fish in the size-composition subsample as juvenile or adult.
3. Based on length-weight relationships, calculate the weight of each fish in the size-composition subsample.
4. Add up the weights of the juveniles and the weights of the adults in the size-composition subsample, and obtain the proportion of the catch made up by each stage.
5. Use those proportions to break down the total catch in the haul into catch of juveniles and catch of adults, and turn to CPUE.
6. Check that it maps to the total CPUE from point 1.

**IMPORTANT**: this approach assumes that the subsample taken for the size-composition is representative of size composition of the entire haul.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(kableExtra)
library(sf)
library(raster)
library(viridis)
library(maps)
library(mapdata)
library(rfishbase)
library(FishLife)
```

```{r}
select <- dplyr::select
```

Read in Atlantis groups, the RACE species list, and map them to one another.
```{r}
atlantis_groups <- read.csv("GOA_Groups.csv", fileEncoding = "UTF-8-BOM")
atlantis_groups <- atlantis_groups %>% select(Code, Name, LongName)

race_species_all <- read.csv("RACE_species_goa_Atlantis_Nov162021.csv", fileEncoding = "UTF-8-BOM")
race_species <- race_species_all[!is.na(race_species_all$Atlantis.group),] # drop NAs (egg cases, debris, etc.)
race_species <- race_species %>% select(Atlantis.group:Scientific.Name)

race_species <- race_species %>% left_join(atlantis_groups, by = c("Atlantis.group" = "Code"))
```

# Catch

Read in AKFIN "Catch by Haul" [data](https://akfinbi.psmfc.org/analytics/saw.dll?Dashboard&PortalPath=%2fshared%2fStock%20Assessment%2f_portal%2fStock%20Assessment&Page=RACE%20-%20Catch%20by%20Haul&Done=Dashboard%26PortalPath%3d%252fshared%252fStock%2520Assessment%252f_portal%252fStock%2520Assessment%26Page%3dRACE%2520Survey%2520%26ViewState%3d16frh4d5qtd0rd5dqoqoctqg12). Data retrieved on November 13 2021. Search options were with all fields left blank except for Survey=GOA. 
```{r}
catch <- read.csv("race_catch_by_haul.csv", skip = 5)

# how many species do we have catch data for, from AKFIN?
length(levels(factor(catch$Species.Code))) # 1560

# and how many hauls?
length(levels(factor(catch$Haul.Join.ID))) # 12827
```

The Catch data does not include information about the effort. So let's use the "Haul Description" [data set](https://akfinbi.psmfc.org/analytics/saw.dll?Dashboard&PortalPath=%2fshared%2fStock%20Assessment%2f_portal%2fStock%20Assessment&Page=RACE%20-%20Haul%20Descriptions&Done=Dashboard%26PortalPath%3d%252fshared%252fStock%2520Assessment%252f_portal%252fStock%2520Assessment%26Page%3dRACE%2520Survey%2520%26ViewState%3df5u8r7769kpetli93l0c71962u) to obtain haul information. The column "Satisfactory Performance" will be used to subset these to the hauls that can be used. Data retrieved in April 2021.
```{r}
hauls_tmp <- read.csv("Haul Descriptions.csv", fileEncoding = "UTF-8-BOM")
hauls <- hauls_tmp %>% 
  select(Haul.Join.ID, Distance.Fished..km., Net.Width..m., Satisfactory.Performance) %>%
  filter(Satisfactory.Performance == "Y") %>% 
  mutate(My.effort.km2 = Distance.Fished..km.* Net.Width..m. * 0.001) %>%
  select(Haul.Join.ID, My.effort.km2)
```

Now join the haul data to the catch data by Haul.Join.ID. Because now the haul data only contains the hauls we can use for effort/CPUE, base the join on that. Drop columns as appropriate. 
```{r}
catch_short <- catch %>% select(Year, Haul.Join.ID, Catch.Join.ID, Ending.Latitude..dd., Ending.Longitude..dd., Bottom.Depth, Species.Code, Scientific.Name, Common.Name, Weight..kg., Number.of.Fish)
```

# Size composition

Read in size composition data.
```{r}
sizecomp <- read.csv('race_length_by_haul.csv', skip=5)

y <- sizecomp %>% filter(Haul.Join.ID==30895) %>% select(Scientific.Name) %>% distinct() %>% pull()
yy <- catch_short %>% filter(Haul.Join.ID==30895) %>% select(Scientific.Name) %>% distinct() %>% pull()
```

How many of the hauls in the catch have size composition information of some kind? Match the hauls in the two datasets
```{r}
h1 <- sizecomp %>% select(Haul.Join.ID) %>% distinct() %>% pull()
h2 <- catch_short %>% select(Haul.Join.ID) %>% distinct() %>% pull()
common_hauls <- intersect(h1,h2)
length(common_hauls)/length(h2) # 0.9689717 of catch hauls
length(common_hauls)/length(h1) # 1 length hauls
```
Some of the hauls in the catch data do not have size composition information (<4%, not a problem).

The main issue actually is that not all species are sampled for size composition in each haul. For example, there may be a catch of pollock in a haul, but that does not mean that pollock was sampled for size composition in that same haul. This seems to be far more pervasive.

Subset size comp data.
```{r}
sizecomp <- sizecomp %>% filter(Haul.Join.ID %in% common_hauls)
```

First of, what species do we have in the size comp set?
```{r}
species <- sizecomp %>% 
  group_by(Common.Name,Scientific.Name,Sex) %>% 
  tally() %>% 
  ungroup() %>%
  arrange(desc(n)) #%>%
  # mutate(prop=n/sum(n)*100) %>%
  # filter(prop>1)
```

Here we need:

1. Length at 50% maturity for each species to assign the correct stage to each row.
2. A length-weight relationship for each species. Ideally these will be given to us by the RACEBASE people. Second choice is to use values provided at the end of von Szalay and Raring (2016). Third is just use the values we got for the parametrization work. Whatever we use, make sure that the units make sense, given that this data set reports length in mm.

## Length at maturity

For the purpose of splitting between juveniles and adults, we use estimates of length at maturity from the `FishLife` [package](https://github.com/James-Thorson-NOAA/FishLife). This is approximate and ideally we will swap out with Alaska-specific values for the groundfish.
```{r}
source('funs_lw_lmat.R')

sizecomp_species <- species %>% select(Scientific.Name) %>% distinct() %>%
  mutate(tt = strsplit(Scientific.Name, ' ')) %>%
  rowwise() %>%
  mutate(Genus = tt[1],
         Species = tt[2]) %>%
  select(Genus, Species) %>%
  drop_na() %>%
  ungroup()

# add pollock, it still goes by its old name in Fishbase for some reason
sizecomp_species <- rbind(sizecomp_species, c('Theragra','chalcogramma'))

fl_params <- (FishLife::FishBase_and_RAM$beta_gv) # all species params
fl_names <- row.names(fl_params)
fl_covmat <- (FishLife::FishBase_and_RAM$Cov_gvv) # all species variance covariance

lmat_key <- sizecomp_species %>% 
  mutate(lmat = purrr::pmap(list(this_genus=Genus,this_species=Species),get_lmat)) %>%
  unnest(cols = c(lmat), keep_empty = T)
```
About 50 species do not get filled with this method. Some of them are because the scientific names do not correspond.

## Length-weight relationships

Taking parameters from [Froese et al. (2014)](https://doi.org/10.1111/jai.12299). The form of the LW relationship in the Bayesian hierarchical analysis is $W=a*L^b$. I apply their code in `SingleSpeciesLWR_7.R` to obtain mean estimates of log10a and b from Bayesian hierarchical analysis based on (1) priors that are specific to body shape of the fish, and (2) previous studies that act as data to get a posterior distribution of the parameters. Importantly, for these parameters, weight is in grams and length is in centimeters.
```{r}
lw_key <- sizecomp_species %>% 
  mutate(lw_bayes = purrr::pmap(list(this_genus=Genus,this_species=Species),possibly(apply_Froese,NA))) %>%
  unnest_wider(col = c(lw_bayes)) %>%
  set_names(c('Genus','Species','Log10a','b')) %>%
  mutate(a=10^Log10a) %>%
  select(-Log10a)
```

Join length-weight and length at maturity keys.
```{r}
lw_lmat <- lw_key %>% 
  left_join(lmat_key, by = c('Genus','Species')) %>%
  drop_na() %>%
  mutate(Scientific.Name = paste(Genus, Species, sep = ' ')) %>%
  select(-Genus,-Species)

# restore pollock to the correct name
lw_lmat$Scientific.Name <- gsub('Theragra chalcogramma','Gadus chalcogrammus',lw_lmat$Scientific.Name)
```

34 species of the original 132 have LW and length at maturity information with this method. That is, for a lot of species we cannot do this. In those instances, assume that the catch is 'adults'. 
```{r}
sizecomp_with_params <- sizecomp %>% left_join(lw_lmat, by = 'Scientific.Name')

# apply length-weight relationship
weight_prop_by_stage <- sizecomp_with_params %>% 
  mutate(Weight.g=a*(Length..mm./10)^b,
         Weight.class.g = Weight.g*Frequency) %>%
  rowwise() %>%
  mutate(Stage=ifelse(is.na(lmat),'U',ifelse(Length..mm./10>lmat,'A','J'))) %>% # undetermined 
  ungroup() %>%
  group_by(Haul.Join.ID,Common.Name,Scientific.Name,Stage) %>%
  summarise(Weight.stage.g=sum(Weight.class.g, na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(Haul.Join.ID,Common.Name,Scientific.Name) %>%
  mutate(Prop.Weight=ifelse(Stage=='U',1,Weight.stage.g/sum(Weight.stage.g))) %>%
  select(-Weight.stage.g)

# change undetermined to adult
weight_prop_by_stage$Stage <- gsub('U','A',weight_prop_by_stage$Stage)
```

Now join the information of the proportion by haul back with the catch data. It becomes apparent here that a lot of hauls do not have size comp information for all species that were caught in the haul. Options:

1. Drop all that is not in the size comp database. This is not going to work, this way we lose a lot of data.
2. Assume that all that is not sized is an adult. Not ideal, but that was the starting point anyway.

Do 2 for now. 
```{r}
catch_short_with_props <- catch_short %>% 
  filter(Haul.Join.ID %in% common_hauls) %>% # subset to the hauls that exist in both the sizecomp and catch data sets
  full_join(weight_prop_by_stage, by = c('Haul.Join.ID','Scientific.Name','Common.Name')) %>%
  rowwise() %>%
  mutate(Stage = ifelse(is.na(Stage),'A',Stage),
         Prop.Weight = ifelse(is.na(Prop.Weight),1,Prop.Weight))
```

Now apportion the catch in weight based on the proportion. 
```{r}
catch_short_with_props <- catch_short_with_props %>%
  mutate(Weight..kg.stage = Weight..kg.*Prop.Weight)
```

Check: does the catch calculated this way add up with the catch in the raw AKFIN data?
```{r}
check_stage <- catch_short_with_props %>% group_by(Haul.Join.ID,Species.Code) %>% summarise(catch=sum(Weight..kg.stage))
check_original <- catch_short %>% group_by(Haul.Join.ID,Species.Code) %>% summarise(catch=sum(Weight..kg.))

check_both <- check_original %>% left_join(check_stage, by=c('Haul.Join.ID','Species.Code')) %>% mutate(diff=catch.x-catch.y)
```
Looking OK.

Then calculate new CPUE for weight and numbers from this. 
```{r}
catch_all <- hauls %>% 
  left_join(catch_short_with_props) %>% 
  mutate(cpue.kg.km2 = Weight..kg.stage/My.effort.km2) %>%
  select(Year, Haul.Join.ID, Catch.Join.ID, Ending.Latitude..dd., Ending.Longitude..dd., Bottom.Depth, Species.Code, Stage, cpue.kg.km2)
```

Some hauls in the haul data do not appear in the catch data. Discard those hauls that do not have information in the catch data. In addition, some records have no CPUE, biomass or numbers. Discard those too.
```{r}
catch_all <- catch_all %>% filter(!is.na(Ending.Latitude..dd.) & !is.na(cpue.kg.km2))
```

Map species to Atlantis groups.
```{r}
# rename column names for consistency with Kirstin's code

catch_all <- catch_all %>% set_names(c("YEAR","HAULJOIN","CATCH","LAT","LON","DEPTHR","CODE","STAGE","BIOM_KGKM2"))

# add a column with Atlantis group code and name
key <- race_species %>% select(Species.Code, Atlantis.group, Name) %>% set_names(c("CODE","ATLANTIS_GROUP","CN"))
catch_all <- catch_all %>% left_join(key, by = "CODE")

catch_all <- catch_all %>% filter(!is.na(CN)) # discard those cases that do not map to Atlantis groups, for now
```

Aggregate species to groups.
```{r}
# sum species to atlantis groups (e.g., catch of different sponge species in the same haul goes toward the "Sponge" group in Atlantis)
cpue <- catch_all %>% group_by(YEAR,HAULJOIN,LAT,LON,DEPTHR,ATLANTIS_GROUP,CN,STAGE) %>% summarise(BIOM_KGKM2 = sum(BIOM_KGKM2)) %>% ungroup()
```

Have a look at the data.
```{r, fig.width = 12, fig.height = 10}
ggplot(data = cpue, aes(x = YEAR))+
  geom_bar(stat = "count")+
  theme_minimal()+
  facet_wrap(~CN, scales = "free_y", ncol = 5)+
  labs(title = "Hauls with non-zero catch")
```

And spatially.
```{r, fig.width = 10, fig.height=16}
cpue_sf <- cpue %>% st_as_sf(coords = c(x = "LON", y = "LAT"), crs = "WGS84")
race_extent <- st_bbox(cpue_sf) # set the extent of the race data

coast <- map("worldHires", regions = c("Canada", "USA"), plot = FALSE, fill = TRUE)
coast <- coast %>% st_as_sf()

group <- "Pollock"
stage <- 'J'

ggplot()+
  geom_sf(data = cpue_sf[cpue_sf$CN == group & cpue_sf$STAGE == stage,], aes(color = log1p(BIOM_KGKM2)))+
  geom_sf(data = coast)+
  coord_sf(xlim = c(race_extent$xmin,race_extent$xmax),ylim = c(race_extent$ymin,race_extent$ymax))+
  scale_color_viridis()+
  theme_minimal()+
  facet_wrap(~YEAR, ncol = 2)+
  labs(title = "Catch by haul")
```
See text to RACE.Rmd for thoughts about sample sizes of each group and what we should use (those refer to individuals, so for groups like corals and sponges there are fewer data points, but the main ideas hold).

Mind, that we are more likely to have convergence issues here. Also, Canada...
```{r}
save(cpue, file = "cpue_by_stage.Rdata")
```
